Viscous Transport in Eroding Porous Media

### 2019-09-24 docs/transport/erosion_transport.tex
working on edits

\todo[inline]{left off here}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry, denoted by $T \geq 1$, is a dimensionless
number that quantifies the average amount of turning made by
trajectories passing through the geometry.  Unlike the for the
dispersion, we do not use reinsertion to form long trajectories.  The
local tortuosity
\begin{align}
  \tau(y) = \frac{\lambda(y)}{d},
  \label{eqn:localTort}
\end{align}
is the length of a single trajectory, $\lambda(y)$, whose initial
$y$-coordinate is a value $y$ on a cross-section of length $d$.  Taking
the average over all points on the cross-section defines the hydraulic
tortuosity
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is a cross-section perpendicular to the $x$-axis with length
$d$, $\lambda(y)$ is the length of the trajectory that is initialized at
the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of the
velocity at the $y$-coordinate along $S$.  Therefore, the tortuosity
satisfies $T \geq 1$, and $T=1$ only if no grains are present.

The definition for the tortuosity depends on how the streamlines are
defined, and Duda et al.~\cite{dud-koz-mat2011} summarize several
options. One choice, that we adopt, considers streamlines that originate
at $S$ and terminate at a parallel cross-section.  We use the
cross-sections $x=-1$ and $x=1$.  This choice of averaging allows us to
also compute the tortuosity in terms of an area integral.  Under the
assumptions that the flow is incompressible and not re-entrant, meaning
that all streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent to~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\cite{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several time steps
in Section~\ref{sec:results}.

There have been efforts to relate the tortuosity to the porosity.  For
example, Matyka et al.~\cite{mat-kha-koz2008} propose the models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $\phi$ is the porosity and $p>0$ is a fitting parameter.  In
Section~\ref{sec:results}, we compare these four models with the
tortuosity of eroding geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in a porous media depends on the porosity, it also
depends on the placement of the grains.  In particular, the grain
placement affects velocity scales~\cite{ali-par-wei-bre2017},
correlation structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019}, and more. In addition,
pore network models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}
rely on descriptions of the pore throat sizes. To characterize the grain
placement, we compute distributions of gap sizes between neighboring
grains.  To define neighboring grains, we form a Delaunay triangulation
of nodes placed at the center of each eroding grain and at a collection
of points around the boundary of the porous region.  Then, we say that
two grains are neighbors if their centers share an edge of the
triangulation~\cite{dea-qua-bir-jua2018}.  Finally, we compute the
minimum distance between all pairs of neighboring grains.  In
Figure~\ref{fig:Eroding100gap}(a), we plot a histogram of the pore size
of 100 eroded bodies.  The black curve is an un-normalized Weibull
distribution with the same mean and variance as the measured data.  In
Figure~\ref{fig:Eroding100gap}(b), we plot lines connected the two
closest points of all neighboring gaps. We do not consider pores between
eroding bodies and the solid and the artificial points placed around the
porous region, so some of the grains near the porous region boundary
only have two neighbors.

\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a): The distribution of the pore
sizes of an eroded geometry. The pores are defined as the space between
neighboring grains.  The black curve is the Weibull distribution with
the same first two moments as the data. (b): Two grains are neighbors if
they share a edge in a Delaunay triangulation of the centers of the
grains.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{sec:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In line with our previous work~\cite{qua-moo2018}, we use two meshes to
simulate erosion. The integral equation is solved by discretizing the
boundary of the geometry at a set of points distributed equally in
arclength, and then applying a collocation method
(Section~\ref{sec:spatialDiscretization}).  Depending on the proximity
of the target point to the source points, the quadrature rule is either
the trapezoid rule or the Barycentric quadrature rule
(Section~\ref{sec:bary}).  The selection algorithm that determines which
quadrature method is applied is described in Section~\ref{sec:fmm}.
Once the shear stress is computed, the bodies are eroded a single time
step by using a {\thL} discretization~\cite{hou-low-she1994}.  By using
a {\thL} formulation, the stiff curvature term is linear, treated
implicitly, and the discretization points remain equispaced in
arclength.  Finally, time stepping methods for the eroding bodies and
passive particles are described in Section~\ref{sec:time}

@@

Once the velocity is computed in $\Omega$, the tortuosity can be
computed with the Eulerian velocity field~\eqref{eqn:tortuosity2}.  We
compute the velocity at $\xx_{ij} = (-1 + i\Delta x, -1 + j\Delta y)$,
$i,j=1,\ldots,N$, where $\Delta x = \Delta y = 2/N$, and the velocity at
points inside an eroding body are assigned a value of 0.  Then, the
tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(\xx_{ij})\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u_x(\xx_{ij}) \Delta x \Delta y}.
\end{align}

@@

Once a collection of trajectories are formed, they are used to quantify
the dispersion and the tortuosity.  The dispersion simply computes the
second-moment of the trajectories as described in
equation~\eqref{eqn:moments}.  We use $N_p = 1000$ streamlines so that
the statistics have converged~\cite{bel-sal-rin1992}.  However, as
described in Section~\ref{sec:dispersion}, a reinsertion method is used
to compute trajectories that are sufficiently long to observe an
asymptotic anomalous dispersion rate.  To compute the tortuosity using
the trajectories, we consider trajectories crossing between the two
cross-sections $x=-1$ and $x=1$.  Then, we discretize
equation~\eqref{eqn:tortuosity1} as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.

@@

In Figure~\ref{fig:Eroding100Transport}(a), we plot the tortuosity as a
function of the porosity.  The initial geometry has a porosity of $\phi
= 50.09\%$ and the tortuosity is $T = 1.20$.  The tortuosity is computed
with both the length of the streamlines~\eqref{eqn:tortuosity1} (red
marks) and using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding100tort}(c).
Again, the two tortuosity formulas give similar results.  For this
geometry, the tortuosity decreases monotonically at almost all
porosities.  However, the tortuosity undergoes a sudden increase near
the end of the simulation.  The increase is caused by a single small
body near the middle of the channel being completely eroded.  While this
results in straighter streamlines, therefore reducing $\lambda$, the
horizontal flow, $u_x(y)$ increases since there is no longer a no-slip
boundary.  At the very late stages of erosion, we have observed that
when a single small body finishes eroding, this can result in a sudden
increase in the tortuosity.  We also computed the lines of best fit
using the porosity-tortuosity models~\eqref{eqn:tortuosityModels}.  The
black dashed line in Figure~\ref{fig:Eroding100Transport}(a) is the line
of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2459}
\end{align}
with a root-mean-square error of $5.50 \times 10^{-3}$.  We note a
slightly better root-mean-square error of $5.20 \times 10^{-3}$ is
possible with the model
\begin{align}
  \widehat{T}(\phi) = 1 - 0.2631 \ln(\phi).
\end{align}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here, the streamline is initialized at the point $(x_0,y_0)$ on a
vertical cross-section of length $d$.  The hydraulic tortuosity is
defined by taking the average over all points on the cross-section
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0}
  {\displaystyle\int_{S}u_1(x_0,y_0)\,dy_0},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is the cross-section $x = x_0$ that is perpendicular to the
$x$-axis with length $d$, and $u_1(y_0)$ is the $x$ component of the
velocity at the initial point of the streamline. Therefore, the
tortuosity satisfies $T \geq 1$, and $T=1$ only if no grains are
present.

The tortuosity definition depends on how the streamlines are
defined, and Duda et al.~\cite{dud-koz-mat2011} summarize several
options. One choice, that we adopt, uses streamlines that originate at
$S$ and terminate at a parallel cross-section.  We use the
cross-sections $x=-1$ and $x=1$ that are at the start and end of the
porous region.  This choice of averaging allows us to also compute the
tortuosity in terms of an area integral.  Assuming that the flow is
incompressible and not re-entrant, meaning that all streamlines connect
the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent to~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_1(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\cite{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in Section~\ref{sec:results}.

There have been efforts to relate the tortuosity to the porosity.  For
example, Matyka et al.~\cite{mat-kha-koz2008} propose the models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $\phi$ is the porosity and $p>0$ is a fitting parameter.  In
Section~\ref{sec:results}, we compare these four models with the
tortuosity of eroding porous geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in a porous media depends on the porosity, it also
depends on the placement of the grains.  In particular, the grain
placement affects velocity scales~\cite{ali-par-wei-bre2017},
correlation structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019}, and more. In addition,
the grain configuration is important for pore network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} that rely on
descriptions of the pore throat sizes. To characterize the grain
placement, we compute distributions of pore sizes between neighboring
grains.  To define neighboring grains, we form a Delaunay triangulation
using nodes placed at the center of each eroding grain and at a collection
of points around the boundary of the porous region.  Then, we say that
two grains are neighbors if their centers share an edge of the
triangulation~\cite{dea-qua-bir-jua2018}.  Finally, we compute the
minimum distance between all pairs of neighboring grains.  In
Figure~\ref{fig:Eroding100gap}(a), we plot a histogram of the pore size
of 100 eroded bodies.  The black curve is an unnormalized Weibull
distribution with the same mean and variance as the measured data.  In
Figure~\ref{fig:Eroding100gap}(b), we plot illustrate the pore throat
sizes by drawing the shortest lines between neighboring grains.  We do
not consider pores between eroding bodies and the solid and the
artificial points placed around the porous region, so some of the grains
near the porous region boundary only have two neighbors.

\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a): The distribution of the pore
sizes of an eroded geometry. The black curve is the Weibull distribution
with the same first two moments as the data. (b): We only consider pore
sizes between neighboring grains, where two grains are neighbors if they
share a edge in a Delaunay triangulation using nodes placed at the
center of each eroding grain.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{sec:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{left off here}
In line with our previous work~\cite{qua-moo2018}, we use two meshes to
simulate erosion. The integral equation is solved by discretizing the
boundary of the geometry at a set of points distributed equally in
arclength, and then applying a collocation method
(Section~\ref{sec:spatialDiscretization}).  Depending on the proximity
of the target point to the source points, the quadrature rule is either
the trapezoid rule or the Barycentric quadrature rule
(Section~\ref{sec:bary}).  The selection algorithm that determines which
quadrature method is applied is described in Section~\ref{sec:fmm}.
Once the shear stress is computed, the bodies are eroded a single time
step by using a {\thL} discretization~\cite{hou-low-she1994}.  By using
a {\thL} formulation, the stiff curvature term is linear, treated
implicitly, and the discretization points remain equispaced in
arclength.  Finally, time stepping methods for the eroding bodies and
passive particles are described in Section~\ref{sec:time}

@@

Once the velocity is computed in $\Omega$, the tortuosity can be
computed with the Eulerian velocity field~\eqref{eqn:tortuosity2}.  We
compute the velocity at $\xx_{ij} = (-1 + i\Delta x, -1 + j\Delta y)$,
$i,j=1,\ldots,N$, where $\Delta x = \Delta y = 2/N$, and the velocity at
points inside an eroding body are assigned a value of 0.  Then, the
tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(\xx_{ij})\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u_1(\xx_{ij}) \Delta x \Delta y}.
\end{align}

@@

Once a collection of trajectories are formed, they are used to quantify
the dispersion and the tortuosity.  The dispersion simply computes the
second-moment of the trajectories as described in
equation~\eqref{eqn:moments}.  We use $N_p = 1000$ streamlines so that
the statistics have converged~\cite{bel-sal-rin1992}.  However, as
described in Section~\ref{sec:dispersion}, a reinsertion method is used
to compute trajectories that are sufficiently long to observe an
asymptotic anomalous dispersion rate.  To compute the tortuosity using
the trajectories, we consider trajectories crossing between the two
cross-sections $x=-1$ and $x=1$.  Then, we discretize
equation~\eqref{eqn:tortuosity1} as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_1(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_1(y_i) \Delta y}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.

@@

In Figure~\ref{fig:Eroding100Transport}(a), we plot the tortuosity as a
function of the porosity.  The initial geometry has a porosity of $\phi
= 50.09\%$ and the tortuosity is $T = 1.20$.  The tortuosity is computed
with both the length of the streamlines~\eqref{eqn:tortuosity1} (red
marks) and using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding100tort}(c).
Again, the two tortuosity formulas give similar results.  For this
geometry, the tortuosity decreases monotonically at almost all
porosities.  However, the tortuosity undergoes a sudden increase near
the end of the simulation.  The increase is caused by a single small
body near the middle of the channel being completely eroded.  While this
results in straighter streamlines, therefore reducing $\lambda$, the
horizontal flow, $u_1(y)$ increases since there is no longer a no-slip
boundary.  At the very late stages of erosion, we have observed that
when a single small body finishes eroding, this can result in a sudden
increase in the tortuosity.  We also computed the lines of best fit
using the porosity-tortuosity models~\eqref{eqn:tortuosityModels}.  The
black dashed line in Figure~\ref{fig:Eroding100Transport}(a) is the line
of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2459}
\end{align}
with a root-mean-square error of $5.50 \times 10^{-3}$.  We note a
slightly better root-mean-square error of $5.20 \times 10^{-3}$ is
possible with the model
\begin{align}
  \widehat{T}(\phi) = 1 - 0.2631 \ln(\phi).
\end{align}
